Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{{DISTRIBUTION STATEMENT A. Approved for public release.} }

### 2021-05-14 paper/agree.tex
Revise agree semantics

\subsection{Meaning of AGREE specifications}

At this point it is important to state the meaning of a high-assurance
component more formally. Assume that the inputs to an AGREE contract
specification are raw strings (e.g., contiguous sequences of bytes)
for each named port of the component, such as what might be sent over
a wire by the communication fabric. An \emph{environment},
$\theta: \lval \mapsto \konst{string}$, divides the raw string into
its constituent parts as given by its data definition in AADL. As
such, it binds \emph{L-values} to strings where an L-value (a concept
from compilers) represents anything that can appear on the left-hand
side of an assignment. In the context of this paper, an $\lval$ is any
reference to a named port, field in a record, entry in an array, local
value defined by an eq-statement etc. The environment $\theta$ maps
any L-value in the scope of the AGREE contract specification to its
associated raw representation as a subset of bytes in a contiguous
string.\footnote{Might want to discuss why we want to stay down at the
representation level.}

[\emph{So basically $\theta$ covers the inputs and the state, i.e. we can
generate a contig type for the inputs plus state vars.} ]

[Question: what is a \emph{total} environment? One that supplies at
least the bindings in $\theta$ needed to evaluate the spec?]

Let $s$ be the AGREE contract specification for a high-assurance
component. The function $\konst{eval}\; s\; \theta\dots$ evaluates $s$
on the given stream of total environments, $\theta\dots$, and
returns \konst{true} if the specification is invariant along the
entire stream and \konst{false} otherwise. The meaning of invariance
for the specification merits some discussion. A guarantee has its
intuitive meaning: the truth value of the associated expression at
every step of the stream must be preserved. An eq-statement, however,
is treated as a guarantee rather than a computation that adds a new
binding to the environment. It asserts an equivalence between the
named value and its associated expression. That equivalence must be
preserved at every step of the stream.

The set $\Theta_s$ is the universe of environment bindings for the
given specification $s$.  The meaning of an AGREE contract
specification is now defined as follows.
%
\[
  \mathcal{L}(s) = \set{\theta\ldots \in \Theta_s^* \mid \konst{eval}\ s\ \theta\ldots}
\]
%
The meaning is the prefix closed set of finite environment streams on
which the specification is invariant. Each environment in the stream
is total, meaning that it binds all L-values in the scope if the
specification.

We claim that a synthesized high-assurance component generates
$\mathcal{L}(s)$ precisely. Let $\konst{synth}(s)$ be the CakeML
synthesized component, and let $\theta^\prime
= \konst{eval}\ \konst{synth}(s)\ \theta\ldots$ be a function that
evaluates the synthesized component on the input stream to produce a
corresponding output stream that is total in its bindings. Let
$I_s(\theta\ldots)$ be a function that maps an environment stream to a
new environment stream that only retains the input bindings of each
environment in the stream. We then claim the following:
%
\[
  \forall \theta\ldots \in \mathcal{L}(s)\ (\konst{eval}\ \konst{synth}(s)\ I_s(\theta\ldots) = \theta\ldots) \\
\]
%

>>>>

\subsection{Brief Semantics Definition}

Here the formal semantics of the AGREE contract specification are briefly
presented to make clear the meaning of a high-assurance component. 
These semantics are used in the next section to argue that the synthesis
preserves the same input to output behavior as the AGREE specification.

Assume that all data is in its raw form which is a contiguous sequence of bytes
representing exactly what is sent over a wire by the communication fabric, so a
datum is given by a \konst{string}. This assumption is important to the 
correctness argument in the next section. An \emph{environment},
$\theta: \lval \mapsto \konst{string}$, binds \emph{L-values} to strings where 
an L-value is anything that can appear on the left-hand side of an assignment
such as a named port, a field in a record, a entry in an array, a local
value defined by an eq-statement etc.

Let $s$ be an AGREE contract specification for some high-assurance
component. The notation $\theta_s$ is used to denote the environment that
contains a binding for any L-value in the scope of $s$ and nothing else, and
the notation $\Theta_s$ denotes the universe of all such environments. The
semantics of $s$ are defined over streams, $\pi = \theta_1, \theta_2, \ldots$, which
are finite sequences of environments, $\pi \in \Theta_s^*$.

\begin{comment}
[\emph{So basically $\theta$ covers the inputs and the state, i.e. we can
generate a contig type for the inputs plus state vars.} ]

[Question: what is a \emph{total} environment? One that supplies at
least the bindings in $\theta$ needed to evaluate the spec?]
\end{comment}

The function $\konst{eval}\; s\; \pi$ is defined such that it evaluates $s$ on
the stream, $\pi$, and returns \konst{true} if $s$ is invariant along the
entire stream and \konst{false} otherwise. A guarantee in $s$ is invariant if
its associated expression is true in the context of every step of the stream, while
an eq-statement in $s$ is invariant if its binding in the context of every step
is equivalent to the computed value of its associated expression in that same
context. All guarantees and eq-statements must be invariant in the stream for
the function to return \konst{true}.

The meaning of an AGREE contract
specification is now defined as the set of environment streams on which it is invariant.
%
\[
  \mathcal{L}(s) = \set{\pi \in \Theta_s^* \mid \konst{eval}\ s\ \pi = \konst{true}}
\]
%
\noindent Intuitively, any stream $\pi \in \mathcal{L}(s)$, at each step, binds
the L-values in the eq-statements
in a way that is consistent with their associated expressions and the guarantees are all
true in that same step.

We claim that a synthesized high-assurance component preserves the input to output
relationship in the specification $s$ over every stream in $\mathcal{L}(s)$. 
Let $\pi^\prime
= \konst{SynthEval}\ s\ \pi$ denate a function that synthesizes $s$ and then
evaluates that synthesized component on the stream $\pi$ to create a new
stream $\pi^\prime$ containing added output and other bindings. We
say that two streams, $\pi_a = \pi_b$ are equivalent in regards to a
specification $s$ if and only if the two streams are the same length and
agree on bindings for the input and output for $s$ at every step of the streams.
We now formally state the correctness claim for synthesis.
%
\[
  \forall \pi \in \mathcal{L}(s)\ ((\konst{SynthEval}\ s\ I_s(\pi)) =_s \pi)\\
\]
%
where $I_s(\pi)$ returns the corresponding stream that only retains bindings
for inputs in each step and nothing else. The claim is that the synthesized component
generates the same output stream as the specification for any stream
belonging to the specification that is restricted to just input binding at each step.
